---
title: "Reproducible Research: Peer Assessment 1"
author: "Aroogz"
date: "September 24, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
```{r echo=FALSE}
options(warn = -1)
```

## Preparing the workspace

load/install the required packages:

```{r message=FALSE}
rm(list = ls())
if("ggplot2" %in% row.names(installed.packages())== FALSE){install.packages("ggplot2")}
if("dplyr" %in% row.names(installed.packages())== FALSE){install.packages("dplyr")}
if("xtable" %in% row.names(installed.packages())== FALSE){install.packages("xtable")}

library(ggplot2)
library(dplyr)
library(xtable)
```

check for file in working directory:

```{r cache=TRUE}
if (!file.exists("activity.csv")){
  
  fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileurl, destfile = "activity.zip")
  unzip("activity.zip", exdir = ".")
}
```
## Loading and preprocessing the data

Load the data:

```{r }
activity <- read.csv(file = "activity.csv", stringsAsFactors = F)
str(activity)
#formatting the date column
activity$date <- as.Date(activity$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

aggregating total and average number of steps:

```{r }
steps_per_day = aggregate(activity["steps"], by= activity["date"], FUN= sum)
ave_steps_per_day = aggregate(activity["steps"], by= activity["date"], FUN= mean)
```
plot histogram of total number of steps taken:

```{r }
ggplot(steps_per_day) + 
            geom_histogram(aes(steps_per_day$steps), 
                           bins = 20, col= "red", 
                           fill= "blue", alpha = 0.2)+
            labs(x = "Total number of steps", 
                          title = "Total Number of Steps Taken Each Day")
```

calculate the mean and median :

```{r showtable, results= "asis"}
step_summary <- summarise(.data= group_by(activity, date), mean_steps = mean(steps), median_steps = median(steps))
xt <- xtable(step_summary)
print(xt, type= "html")
```

## What is the average daily activity pattern?

calculate number of steps for different intervals averaging across the days:

```{r }
step_periods <- summarise(.data = group_by(activity, interval), step_ave_across_days = mean(steps, na.rm = T))
```

Time series plot of average number of steps taken across the days:

```{r }
ggplot(step_periods, aes(interval, step_ave_across_days))+geom_line() +labs(x = "Time Interval Along the Day", y = "Average No. of Steps Across all Days", title= "Average Daily Pattern")
```

5-minute interval with the maximum number of steps, averaging over days:

```{r}
step_periods$interval[which.max(step_periods$step_ave_across_days)]
```

## Imputing missing values

calculate and report the total number of missing values in the dataset:

```{r}
sum(!complete.cases(activity))
```

imputing missing values for missing data with `replacement`:

```{r}
na_positions <- which(is.na(activity), arr.ind = T)
loc <- as.data.frame(na_positions)
rows <- loc$row
replacement <- vector()

for (i in seq(nrow(na_positions))){
  
  interval <- activity$interval[rows[i]]
  row_in_step_periods <- which(step_periods$interval == interval)
  replacement <- c(replacement, step_periods$step_ave_across_days[row_in_step_periods])
}
```

create a new dataset:

```{r}
new_activity <- activity
new_activity[is.na(new_activity)] <- round(replacement)
```

calculate the mean and median:

```{r}
new_steps_per_day_summary <- summarise(group_by(new_activity, date), 
                               total_step = sum(steps), 
                               mean_step = mean(steps), 
                               median_steps = median(steps))
```

plot histogram:

```{r }
ggplot(new_steps_per_day_summary)+
            geom_histogram(aes(new_steps_per_day_summary$total_step), 
                           bins = 20, col= "red", 
                           fill= "blue", alpha = 0.2)+
  labs(x = "Total number of steps", 
       title = "Total Number of Steps Taken Each Day (with filled data)")
```

## Are there differences in activity patterns between weekdays and weekends?

create factor variable `weekday` specifyin whether date is weekend or weekday :

```{r}
week_days <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
new_activity$weekday <- factor((weekdays(new_activity$date) %in% week_days), 
                               levels = c(TRUE, FALSE), labels = c("weekday", "weekend"))
```

aggregating the new data set over the weekday variable and interval:

```{r}
new_step_period <- summarise(group_by(new_activity, weekday, interval), 
                             interval_average = mean(steps))

```

plot showing the Average number of steps for weekends and weekeds:

```{r }
ggplot(new_step_period) + geom_line(aes(interval, interval_average, col= "red"))+
                          facet_grid(weekday~.)+
                          labs(x= "Time Interval Along the Day", y= "Average No. of Steps Across all Days",
                               title= "Average Daily Pattern weekend/weekday")

```

Thanks for reading.

&copy; Aroge

